Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-11-20 inst/doc/nonparametric-bayes.Rmd
sensitivity analysis text added, code purled

```{r }
allen_priors_xtable <- data.frame(parameter = c("r0", "K", "theta", "sigma"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r}
ricker_priors_xtable <- data.frame(
  parameter = c("r0", "K", "sigma"),
  "lower bound" = c(r0_prior_p[1], K_prior_p[1], stdQ_prior_p[1]),
  "upper bound" = c(r0_prior_p[2], K_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r } 
myers_priors_xtable <- data.frame(parameter = c("r0", "K", "theta", "sigma"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE, cache=FALSE}
xtable::xtable(rbind(dictable, aictable, bictable))
```

@@

```{r Figure4, dependson=c("profits"), fig.cap="Histograms of the realized net present value of the fishery over a range of simulated data and resulting parameter estimates. For each data set, the three models are estimated as described above. Values plotted are the averages of a given policy over 100 replicate simulations. Details and code provided in the supplement.",  include=TRUE, echo=FALSE}

```

@@

Appendix
========

Model definitions and estimation
---------------------------------

@@

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } \label{rickereq} $$

@@

@@

@@

Eq S3: Allen model

@@

Figure S3: Allen model: prior and posterior distributions for parameter estimates.

@@

## Optimal Control Problem

@@

### Training data

Eacho of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = $ ` r p[1]`, 
$K =$ ` r p[2]`, $C =$ ` r p[3]`, and  $\sigma_g =$ ` r sigma_g` 
for $T=$ `r Tobs` timesteps, starting at initial condition $X_0 = $ `r Xo`. 
The training data can be seen in Figure 1.

### Sensitivity Analysis

>>>>

```{r allen_priors_xtable, include=TRUE, echo=FALSE}
allen_priors_xtable <- data.frame(parameter = c("r0", "K", "theta", "sigma"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r ricker_priors_xtable, include=TRUE, echo=FALSE, cache=FALSE}
ricker_priors_xtable <- data.frame(
  parameter = c("r0", "K", "sigma"),
  "lower bound" = c(r0_prior_p[1], K_prior_p[1], stdQ_prior_p[1]),
  "upper bound" = c(r0_prior_p[2], K_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r myers_prior_xtable, include=TRUE, echo=FALSE} 
myers_priors_xtable <- data.frame(parameter = c("r0", "K", "theta", "sigma"),
           "lower bound" = c(r0_prior_p[1], K_prior_p[1], theta_prior_p[1], stdQ_prior_p[1]),
           "upper bound" = c(r0_prior_p[2], K_prior_p[2], theta_prior_p[2], stdQ_prior_p[2]))
```

@@

```{r Table1, dependson="deviances", include=TRUE, results="asis", echo=FALSE, cache=FALSE}
xtable::xtable(rbind(dictable, aictable, bictable), caption="Model choice scores for several common criteria all (wrongly) select the simplest model. As the true (Allen) model is not distinguishable from the simpler (Ricker) model in the region of the observed data, this error cannot be avoided regardless of the model choice criterion. This highlights the danger of model choice when the selected model will be used outside of the observed range of the data.")
```

@@

```{r Figure4, dependson=c("profits"), fig.cap="Histograms of the realized net present value of the fishery over a range of simulated data and resulting parameter estimates. For each data set, the three models are estimated as described above. Values plotted are the averages of a given policy over 100 replicate simulations. Details and code provided in the supplement.",  include=TRUE, echo=FALSE, fig.width=6, fig.height=6}

ggplot(actual_over_optimal, aes(value)) + geom_histogram(aes(fill=variable)) + 
  facet_wrap(~variable, ncol=2)  + 
  guides(legend.position = "none") +
  xlab("Total profit by replicate") + 
  scale_fill_manual(values=colorkey) # density plots fail when delta fn

# ggplot(actual_over_optimal, aes(value)) + geom_histogram(aes(fill=variable), binwidth=0.1) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)
# ggplot(actual_over_optimal, aes(value, fill=variable, color=variable)) + # density plots fail when delta fn
#  stat_density(aes(y=..density..), position="stack", adjust=3, alpha=.9) + 
#  xlab("Total profit by replicate")+ scale_fill_manual(values=colorkey)+ scale_color_manual(values=colorkey)
```

Sensitivity Analysis
--------------------

These results are not sensitive to the modeling details of the simulation.  The BNP-SDP estimate remains very close to the optimal solution obtained by knowning the true model across changes to the training simulation, noise scale, parameters or structure of the underlying model, as seen in the figure. Results are pooled across different random seeds (111, 222, 333), noise values of 0.01 and 0.05 (not large enough to violate the self-sustaining criterion of @Reed1979), and different randomly generated parameters sets for each model, and using either the Myers or Allen as the underlying structure.  While the BNP-SDP performs slightly closer to optimal on the smaller noise levels, and closer on the Allen model then the Myers model, the differences are small compared to the values observed in using the wrong parameteric model.  The BNP-SDP solution consistently avoids either failing to harvest at all or crashing the stock, which results in the very low profits/values seen in parameteric models in the previous figure.

![Sensitivity Analysis](figure/sensitivity.pdf)

@@

Model definitions and estimation
=================================

Ricker Model
------------

@@

The Ricker model is given by

(@rickereq) $$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } $$

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2Jags] (see provided code).

@@

Myers Model
-----------

The Myers model [@Myers1995] is given by

@@

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, $\theta$ controls the strength of the nonlinearity, exhibiting an allee effect for $\theta \geq 2$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2Jags] (see code provided).

@@

Allen model
-----------

The Allen model [@Allen2005] is given by

@@

where $Z_t$ is log-normal noise of mean unity and log standard deviation $\sigma$, representing the stochastic growth, $X_t$ the stock size at time $t$, $S_t$ the escapement (unharvested population that will recruit in the following year, $S_t = X_t - h_t$).  We place uniform priors on the growth rate $r$, carrying capacity $K$, allee threshold $C$, and log-normal standard deviation parameter $\sigma$, over ranges given in Table 1. Posteriors are inferred by Gibbs sampling using Jags [@R2Jags] (see code provided).

<!-- Figure S3: Allen model: prior and posterior distributions for parameter estimates. -->

@@

Optimal Control Problem Specification 
-------------------------------------

@@

Training data
-------------

Eacho of our models $f(S_t)$ must be estimated from training data, which
we simulate from the Allen model with parameters $r = $ `r p[1]`, 
$K =$ `r p[2]`, $C =$ `r p[3]`, and  $\sigma_g =$ `r sigma_g` 
for $T=$ `r Tobs` timesteps, starting at initial condition $X_0 = $ `r Xo`. 
The training data can be seen in Figure 1.

Code
----

A copy of the script to reconstruct the simulations and analysis shown here is provided in the supplemental materials, and through this version-stable link to the project's Github code repostitory, [nonparameteric-bayes.R]().  This code is dynamically embedded into the manuscript using Knitr, [@knitr].  The script relies on custom routines for executing the estimation of the Gaussian process and the for solving the stochastic dynamic programming problem.  The sensitivity analysis is provided as a separate script, [sensitivity.Rmd](). These routines are provided as an R package, [nonparameteric-bayes](), also available on Github.

Data
----

While the data can be regenerated using the code provided, for convenience CSV files of the data shown in each graph are also provided, along with appropriate metadata written in the Ecological Metadata Language (EML).
